---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# GTFS2EMIS

Respository of the R package to estimate public transport emissions based on GTFS files.

### Installation

```{r,message=FALSE}
devtools::install_github("rafapereirabr/gtfs2emis",auth_token = "94c88a642a2f62fce15e4c05cc0b6ead55ff5851")
```
### Usage

For now, the function are a little bit adapted to Curitiba's GTFS and fleet characteristics. Also, it still possible to 
simulate emissions only for CO and NOx.

```{r}
# ef_hdv_speed(vel = units::set_units(10,km/h),veh = "Urban Buses Midi <=15 t",
#             fuel = "Diesel",euro = "Euro III",tech = NA,pol = "CO",show.equation = TRUE)
#
# gtfs2gps(raw_gtfs = "inst/extdata/gtfs_cur.zip",filepath = "test_joao/gps/",filter_weekdays = TRUE)
#
# gps_to_linestring(input_filepath = "test_joao/gps/",output_filepath = "test_joao/lines/",
#                  fleet_data = "inst/extdata/cur_fleet.tar.xz",overwrite = FALSE)
```

### Preparation scripts

##### 01) `prep/01_*.R` - prep scripts 

1 - download gps cur

2 - prep fleet cur

3 - check valid shapeids (perhaps it could be add into gtfs2gps package)

4 - create hexagons (not working so far I dont know why)

5 - read_gps (from points data to linestring)

##### 02) `prep/02_*.R` - emission

1 - emission estimation

##### 03) `prep/03_*.R` - main 

1 - main script

##### 04) `prep/08_*.R` - plot and visualization

used only for check data purposes

##### 05) `prep/09_*.R` - auxiliar script / internal functions

1 - emission factor from EMEEA

### Project

The main steps of the development should be presented in `projects` sheet