---
title: 'Introduction to gtfs2emis'
author: "Joao Bazzo, Rafael H. M. Pereira, Pedro R. Andrade"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "`gtfs2emis` is an R package to estimate public transport emissions based on GTFS data. The package allows users to estimate the emission levels of several types of pollutants at high spatial and temporal resolutions simply using a GTFS feed and some information on fleet characteristics."
urlcolor: blue
vignette: |
  %\VignetteIndexEntry{Introduction to gtfs2emis} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

# Introduction

`gtfs2emis` is an R package to estimate hot exhaust emissions of public transport systems based on GTFS data. The package allows users to estimate the emission levels of several types of pollutants at high spatial and temporal resolutions simply using a GTFS feed and some information on fleet characteristics. This vignette introduces the main functions of the `gtfs2emis` package.


# Installation

You can install `gtfs2emis`:

```{r, eval = FALSE, message = FALSE}
# From CRAN
install.packages("gtfs2emis")

# or use the development version with latest features
utils::remove.packages('gtfs2emis')
devtools::install_github("ipeaGIT/gtfs2emis")

```

# Package overview

Before we start, let's load packages we'll be using in this vignette:

```{r, message = FALSE}
library(gtfs2emis)
library(gtfstools)
library(gtfs2gps)
library(data.table)
library(sf)
library(units)
library(magrittr)
library(ggplot2)
```


The `gtfs2mis` has a few core functions:

* Estimate public transport emissions
  - `transport_model()` to convert GTFS data into GPS-like `data.table` with the space-time positions and speeds of public transport vehicles.
  - `emission_model()` to estimate hot-exhaust emissions based on (input from the transport model and data on fleet characteristics passed by the user.
  
* Data post-processing

- `emis_summary()` to aggregate emission estimates by time, vehicle type or road segment.
- `emis_grid()` to spatially aggregate emission estimates by using any custom spatial grid.



# Demonstration on sample data

To demonstrate how the `gtfs2emis` package works, we will be using a small sample of data for the city of Porto Alegre (Brazil). Let's load the data.

In this case, we'll only be estimating the emissions of service during business days.


<!-- The first step selects the GTFS and the desired route to model emissions. Here we use one example of `shape_id == "176-1"` in Porto Alegre (Brazil). Detailed functions of GTFS editing can be found in `gtfs2gps` and `gtfstools` packages. -->

```{r, message = FALSE}
gtfs <- gtfs2gps::read_gtfs(system.file("extdata/poa.zip", package = "gtfs2gps")) %>%
        # gtfstools::filter_by_weekday(weekday ='monday') 
        gtfs2gps::filter_week_days() 
        # %>% gtfstools::filter_by_shape_id(shape_id = '176-1') 



```


## 1. Transport model

The first step is to generate the transport model using the `transport_model()` function. This function converts GTFS data into a GPS-like data.table, fixing extreme or missing values of speed and transforming the output into a sf-linestring, which is the required input for emissions estimates. The user can input either a string with the file path where the a gtfs.zip file is stored, or an object of class `"gtfs" "list"`, generated with `gtfs2gps::read_gtfs()`.

# [ RAFA: parei aqui]

```{r, message = FALSE}
sf_line <- transport_model(gtfs = gtfs, parallel = TRUE)
```
Visualizing the transport model output for `shape_id == "176-1"` of Porto Alegre (Brazil).
```{r, message = FALSE, fig.width=3,fig.height=3}
head(sf_line$geometry)
sf_line <- sf::st_as_sf(sf_line)
plot(sf_line["trip_number"])
```




- It considers the fleet to be homogeneously distributed across the vehicle routes.

**3. Fleet data**\
After generating the transport model, we advance to estimate the hot exhaust emissions.
We create a data.frame with fleet information, which can also be imported from an external file. There are two main key features of fleet data:

- If the user has fleet data but does not have information on the relations between buses and routes (e.g Articulated buses to `shape_id` `XX`, `YY` or `ZZ`), she can assign a probability of each fleet type into the routes.  If the class of a given route is known, the probability is equals to 1, always assigning the same class to the route. The sum of all probabilities of each route must be one.
- The fleet data needs to be organized according to the desired emission factor database. For instance, if the user wants to estimate emissions based on typical European fleet, she must present the data according to the typical emission factors inputs for Emep-EEA methodology. In this example, the data is organized according to the requirements for `ef_emep_europe()` emission factor data. Users can check in `ef_moves_usa()`, `ef_emfac_usa()`, `ef_cetesb_brazil()` or `ef_emep_europe()` how the fleet must be organized.

In the example of European emission factors, we use three vehicles classes. A given `shape_id` has 40\%, 50\%, and 10\% chance of having `Ubus Midi <=15 t`, `Ubus Std 15 - 18 t`, and `Ubus Artic >18 t`, respectively.  

```{r, message = FALSE}
fleet_data_ef_europe <- data.table::data.table("veh_type" = c("Ubus Midi <=15 t"
                                                              ,"Ubus Std 15 - 18 t"
                                                              ,"Ubus Artic >18 t")
                                               ,"euro" = c("III","IV","V")
                                               ,"fuel" = rep("D",3)
                                               ,"tech" = c("-","SCR","SCR")
                                               ,"fleet_composition" = c(0.4,0.5,0.1))
fleet_data_ef_europe
```
This data.table is quite similar to other emissions factors database sources. Three other examples are described below.

Emission Factor Model (California Air Resources Board):

```{r, message = FALSE}
fleet_data_ef_emfac <- data.table::data.table("veh_type" = "BUS_URBAN_D"
                                              ,"model_year" = 2011:2015
                                              ,"fuel" = "D"
                                              ,"calendar_year" = 2019
                                              ,"fleet_composition" = rep(0.2,5))
fleet_data_ef_emfac
```
 MOtor Vehicle Emission Simulator (MOVES) from Environmental Protection Agency (United States):
```{r, message = FALSE}
fleet_data_ef_moves <- data.table::data.table("veh_type" = "BUS_URBAN_D"
                                              ,"model_year" = 2011:2015
                                              ,"fuel" = "D"
                                              ,"calendar_year" = 2016
                                              ,"fleet_composition" = rep(0.2,5))
fleet_data_ef_moves
```

Environmental Company of São Paulo (CETESB, Brazil):
```{r, message = FALSE}
fleet_data_ef_cetesb <- data.table::data.table("veh_type" = c("BUS_MICRO_D"
                                                              ,"BUS_URBAN_D"
                                                              ,"BUS_ARTIC_D")
                                               ,"model_year" = rep(2010, 3)
                                               ,"fuel" = rep("D", 3)
                                               ,"fleet_composition" = c(0.4, 0.4, 0.2))
fleet_data_ef_cetesb
```

**4. Emissions model**\
To generate the emission results, we couple the results from `transport_model()` along with the `fleet_data` previously imported. The `emission_model` generates a list of several files related to emissions, storing information such as vehicle variables (`fuel`, `age`, `tech`, `euro`, `fleet_composition`), travel variables (`slope`, `load`, `gps`) or pollution (`EF`, `emi`).

```{r, message = FALSE}
sf_emis <- emission_model(gps = sf_line
                          ,ef_data_base = "emep_europe"
                          ,fleet_data = fleet_data_ef_europe
                          ,pollutant = c("CO","PM10","CO2","CH4","NOx"))
class(sf_emis)
names(sf_emis)
```

**5. Data post-processing**\
As the results are stored in a long list, gtfs2emis package has several function to post-process the results. `emis_summary` summarizes the emissions according to the variable of interest, such as vehicle type `veh_type`, `pollutant`, or `time`. In the last case, we need to add a `time_column` in the list, as this information is not returned by `emission_model()`.

```{r, message = FALSE}
sf_emis$time_column <- sf_emis$gps$timestamp

names(sf_emis)
my_emis_summary <- emis_summary(emi_list = sf_emis,
                    emi_var = "emi", 
                    by = "time", 
                    time_column = "time_column",
                    veh_var = "veh_type", 
                    pol_var = "pollutant") 
my_emis_summary
```
To visualize results by vehicle type `veh_type`:
```{r, message = FALSE}
my_emis_summary <- emis_summary(emi_list = sf_emis,
                                 emi_var = "emi", 
                                 by = "veh_type", 
                                 time_column = "time_column",
                                 veh_var = "veh_type", 
                                 pol_var = "pollutant") 
my_emis_summary
```
If available in `names(sf_emis)`, it is possible to add extra argument about fleet.
```{r, message = FALSE}
my_emis_summary <- emis_summary(emi_list = sf_emis,
                                emi_var = "emi", 
                                by = "veh_type", 
                                time_column = "time_column",
                                veh_var = c("veh_type", "euro", "fuel"),
                                pol_var = "pollutant") 
my_emis_summary
```
Checking the results by pollutant:
```{r, message = FALSE}
my_emis_summary <- emis_summary(emi_list = sf_emis,
                                 emi_var = "emi", 
                                 by = "pollutant", 
                                 time_column = "time_column",
                                 veh_var = "veh_type", 
                                 pol_var = "pollutant") 
my_emis_summary
```
If the user wants to have all results in a data.table format:
```{r, message = FALSE}
sf_emis$time_column <- sf_emis$gps$timestamp
my_emis_dt <- emi_to_dt(emi_list = sf_emis
                    ,emi_vars = "emi"
                    ,veh_vars = c("veh_type", "euro", "fuel")
                    ,pol_vars = "pollutant"
                    ,segment_vars = "time_column")
head(my_emis_dt, 5)
```

To visualize emissions spatially in a grid, the user nees to provide info about the desired grid. An operation of intersection can be done using `emis_grid()`. Here, we use `vein::make_grid()` to produce a basic grid.

```{r, message = FALSE}
grid_gps <- vein::make_grid(spobj = sf_emis$gps, width =  0.25 / 102.47) # 500 meters
grid_gps <- sf::st_as_sf(grid_gps)
```

```{r, message = FALSE, fig.width=4,fig.height=4}
names(sf_emis$emi)
# we add info of emissions into gps file
sf_emis$gps <- cbind(sf_emis$gps,sf_emis$emi)

pol_grid <- emis_grid(data = sf_emis$gps,
                     emi = "CO_Euro_IV",
                     grid = grid_gps,
                     time_class = 'all periods')
plot(pol_grid["CO_Euro_IV"])
```







 
**4. Emissions estimates**

As the emissions varies with `shape_id` and `fleet`, we compute emissions separately, in a loop.

```{r}
linestring_poa$shape_id %>% unique()
```

```{r, message = FALSE}
emi_data <- lapply(unique(gps_poa$shape_id),function(i){
  tmp_ef_euro <- ef_emep_europe(speed = linestring_poa[shape_id == i]$speed,
                                ,veh_type = fleet_data_ef_europe[shape_id == i]$veh_type
                                ,euro =  fleet_data_ef_europe[shape_id == i]$euro
                                ,pollutant = c("CO2","NOx")
                                ,fuel =  fleet_data_ef_europe[shape_id == i]$fuel
                                ,tech =  fleet_data_ef_europe[shape_id == i]$tech
                                ,as_list = TRUE)
  tmp_emis <- emis(fleet_composition = fleet_data_ef_europe[shape_id == i]$fleet_composition
                   ,dist = units::set_units(linestring_poa[shape_id == i]$dist,"km")
                   ,ef = tmp_ef_euro
                   ,aggregate = FALSE
                   ,as_list = TRUE)
  return(tmp_emis)
})
```

**5. Emission into grid**

We now show the example for `CO2` emissions.

```{r}
# Create Grid
grid_gps <- vein::make_grid(spobj = sf::st_as_sf(linestring_poa$geometry)
                            , width =  0.25 / 102.47) # 500 meters
grid_gps <- sf::st_as_sf(grid_gps)

tmp_emis_dt <- lapply(1:length(emi_data),function(i){ # i = 3
  tmp_emi_dt <- emi_to_dt(emi_list = emi_data[[i]]
                          ,emi_vars = "emi"
                          ,veh_vars = "veh_type"
                          ,pol_vars = "pollutant")
  tmp_gps <- cbind(linestring_poa[shape_id == unique(gps_poa$shape_id)[i]]
                   ,tmp_emi_dt[pollutant == "CO2"])
  tmp_gps <- sf::st_as_sf(tmp_gps)
  tmp_emi_grid <- emis_grid(data = tmp_gps
                            ,emi = "emi"
                            ,grid = grid_gps)
  return(tmp_emi_grid)
}) %>% data.table::rbindlist()
```

Summing all emissions and plotting:

```{r, fig.width=4,fig.height=4}
tmp_emis_dt <- tmp_emis_dt[,total_emi := sum(emi), by = id][, .SD[1], by = id]
tmp_emis_dt2 <- sf::st_as_sf(tmp_emis_dt)

plot(tmp_emis_dt2["emi"])
```

## Credits

The gtfs2gps package is developed by a team at the Institute for Applied Economic Research (Ipea) with collaboration from the National Institute for Space Research (INPE), both from Brazil. 
If you have any suggestions or want to report an error, please visit the package GitHub page (add link).
<!-- ## End ---- -->