---
title: 'Introduction to gtfs2emis'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "`gtfs2emis` is an R package to estimate public transport emissions using data in the General Transit Feed Specification (GTFS) format. The package allows users to estimate the emission levels of several types of pollutants for each public transport vehicle/trip/route at high spatial and temporal resolutions simply using a GTFS feed and some information on fleet characteristics."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Introduction to gtfs2emis} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

# 1. Introduction

`gtfs2emis` is an R package to estimate hot exhaust emissions of public transport systems using data in the General Transit Feed Specification (GTFS) format. The package allows users to estimate the emission levels of several types of pollutants for each public transport vehicle/trip/route at high spatial and temporal resolutions simply using a GTFS feed and some information on fleet characteristics. This vignette introduces the main functions of the `gtfs2emis` package and shows a step-by-step reproducible example of how to use the package.


# 2. Installation

One can install `gtfs2emis` from CRAN (recommended) or the development version from Github:

```{r, eval = FALSE, message = FALSE}
# From CRAN
install.packages("gtfs2emis")

# Dev. version with latest features
utils::remove.packages('gtfs2emis')
devtools::install_github("ipeaGIT/gtfs2emis")

```


# 3. Data requirements

To estimate the emission levels from a given public transport system, users need to:

1.  Input a `GTFS.zip` file
2.  Input `data.frame` with a few characteristics of the public transport fleet (such as age and fuel)
3.  Select an emission factor model from the models provided by the package, which currently includes models from the US, Europe and Brazil.
4.  Select which pollutants should be estimated from a list with of over 15 pollutants provided by the package.




# 4. Package overview

Before we start, let's load a few packages we'll be using in this vignette:

```{r, message = FALSE}
library(gtfs2emis)
library(gtfstools)
library(progressr)
library(data.table)
library(magrittr)
library(ggplot2)
library(units)
library(sf)

```

The `gtfs2mis` package two core functions:

1. `transport_model()`. This function converts GTFS data into a GPS-like `data.table` with the space-time positions and speeds of public transport vehicles.
2. `emission_model()`. This function estimates hot-exhaust emissions based on four inputs:     
    a) the result from the `transport_model()`;
    b) a `string` indicating which emission factor model should be considered;
    c) a `string` indicating which pollutants should be estimated; and
    d) a `data.frame` with info on fleet characteristics passed by the user. 
    The function returns a `list` with the estimated amount of each pollutant emitted by public transport vehicles.

To  help users analyze the output from `emission_model()`, the `gtfs2emis` package has few functions:

3. `emis_summary()` to aggregate emission estimates by time of the day, vehicle type or road segment.
4. `emis_grid()` to spatially aggregate emission estimates using any custom spatial grid or polygons.
5. `emis_to_dt()` to convert the output of `emission_model()` from `list` to `data.table`.
    


# 5. Demonstration on sample data

In this introductory vignette, we show a very simple case study using default parameters of the `gtfs2emis` package and where we assume that fleet characteristics are homogeneously distributed across the public transport routes. [For advanced users, we have written another vignette](add_link) that demonstrates how `gtfs2emis` allows one to allocate specific vehicles into predefined routes, fix speeds and boarding times in the transport model, and adjust emissions factors. We have also written a separate vignette to help users build the `data.frame` with information on [fleet characteristics](add_link).


To demonstrate how the `gtfs2emis` package works, we will be using a small sample data for the city of Dublin, Ireland. In this example, we'll be estimating `CO2` and `PM10` emissions of bus services on business days.


## 5.1 Transport model

The first step is to generate the transport model using `transport_model{gtfs2emis}`. This function converts GTFS data into a GPS-like `data.table`, transforming the output into a `sf-linestring`, which is the required input for emissions estimates. The user can input either a string with the file path where the a `gtfs.zip` file is stored, or an object of class `"gtfs" "list"`, generated with `gtfstools::read_gtfs()`.

First, let's read the GTFS data and filter only the transport services that run on

```{r, message = FALSE}
# path to GTFS.zip file
gtfs_file <- system.file("extdata/irl_dub/irl_dub_gtfs.zip", package = "gtfs2emis")

# read GTFS
gtfs <- gtfstools::read_gtfs(gtfs_file)

# Keep Monday services GTFS
gtfs <- gtfstools::filter_by_weekday(gtfs, 
                                         weekday = c('saturday', 'sunday'), 
                                         keep = FALSE)

# 66666666666 REMOVER
gtfs <- gtfstools::filter_by_trip_id(gtfs, trip_id = '619.3.60-40-d12-1.224.O')

```


Now let's generate the transport model. This is the most time consuming part. If you want to, you can set a progress bar by calling the `transport_model()` function within `progressr::with_progress()`, as shown below.


```{r, message = FALSE}
# generate transport model
progressr::with_progress( 
  
  tp_model <- transport_model(gtfs_data = gtfs,
                              spatial_resolution = 100,
                              parallel = TRUE) 
  )

head(tp_model)
```

Here is how the output of the transport model looks like. In essence, it's a trajectory `data.table sf linestring` with the space-time position and speed of trips segments for each single vehicle of the public transport system.

```{r, message = FALSE, fig.width=5, fig.height=5}
ggplot(data = tp_model) +
  geom_sf(aes(color= as.numeric(speed))) +
  scale_color_continuous(type = "viridis")+
  labs(color = "Speed (km/h)")+
  theme_void()

```

## 5.2 Fleet data

The next step is to prepare a `data.frame` with some characteristics of the public transport fleet. This can be either: 
- A simple table with the overall composition of the fleet. In this case, the `gtfs2emis` will assume that fleet is homogeneously distributed across all routes; OR
- A detailed table that (1) brings info on the characteristics of each individual vehicle and, (2) tells the probability with which each vehicle type is allocated to each transport route.


In this introductory vignette, we'll be working with a simple table that tell us the proportion of buses in Dublin according to their vehicle type, Euro standard, technology and fuel. The table looks like this:

```{r, message = FALSE}
fleet_file <- system.file("extdata/irl_dub/irl_dub_fleet.txt", package = "gtfs2emis")

fleet_df <- read.csv(fleet_file)
head(fleet_df)

```


Please note different emission factor models may require different information and that the fleet `data.frame` needs to be organized accordingly. In our current example for the city of Dublin, the fleet data must include certain columns with the fleet characteristics that are used in the Emep-EEA emission factor model: vehicle type, Euro standard, technology and fuel. To check which columns and sets of vehicle characteristics are required by a given emission factor model, check the [fleet data vignette](add_link).



## 5.3 Emission model

In the final step, we use the `emission_model{gtfs2emis}` function to estimate hot exhaust emissions of our public transport system. Here, the user needs to pass the results from `transport_model()`, some fleet data as described above, and select which emission factor model and pollutants should be considered. The `emission_model()` outputs a `list` with several `vectors` and `data.frames` with emission estimates and related information such as vehicle variables (`fuel`, `age`, `tech`, `euro`, `fleet_composition`), travel variables (`slope`, `load`, `gps`) or pollution (`EF`, `emi`). 

[BAZZO 666] acho q essa ultima frase texto pode ficar mais clara.

```{r, message = FALSE}
emi_list <- emission_model(tp_model = tp_model
                          , ef_model = "ef_europe_emep"
                          , fleet_data = fleet_df
                          , pollutant = c("CO2","PM10")
                          )

names(emi_list)
```



# 6 Analyzing emission estimates

Sometimes, working with data stored in a long list can be tricky. For convenience, `gtfs2emis` package has a few functions to users analyze the these results and make. For example, you can use the `emis_to_dt()` function to convert the output of `emission_model()` from a `list` to a `data.table` format:

```{r,eval = TRUE}
emi_dt <- emis_to_dt(emi_list = emi_list
                    ,emi_vars = "emi"
                    ,veh_vars = c("veh_type", "euro", "fuel")
                    ,pol_vars = "pollutant")

head(emi_dt) 

```


Alternatively, the `emis_summary()` function helps users summarize emission estimates, aggregating emissions by either `veh_type`, `pollutant`, or `time`:

## 6.1 Total emissions by vehicle type

The `emis_summary()` function help users summarize emission estimates by either `veh_type`, `pollutant`, or `time`. Using this function, users can easily:


```{r, fig.height=5, fig.width=8}
emi_total <- emis_summary(emi_list = emi_list,
                          emi_var = "emi",
                          by = "veh_type",
                          veh_var = "veh_type",
                          pol_var = "pollutant") 
head(emi_total)

# plot
ggplot(data = emi_total) +
  geom_col(aes(x = veh_type, y = as.numeric(emi/1000), fill = pollutant), 
           color=NA, show.legend = FALSE) +
  labs(y="Total emissions (Kg)", x="vehicle type") +
  facet_wrap(~pollutant, scales = "free", nrow = 2) +
  theme_minimal()

```



## 6.2 Total emissions by pollutant type


```{r, message = FALSE}
emi_by_pol <- emis_summary(emi_list = emi_list,
                                 emi_var = "emi", 
                                 by = "pollutant", 
                                 veh_var = "veh_type", 
                                 pol_var = "pollutant") 
emi_by_pol

```


It is also possible to summarize emission estimates by other vehicle characteristics AND pollutant type.

```{r, message = FALSE}
emi_by_veh <- emis_summary(emi_list = emi_list,
                                emi_var = "emi", 
                                by = "veh_type", 
                                veh_var = c("veh_type", "euro", "fuel"),
                                pol_var = "pollutant")

head(emi_by_veh)
```


## 6.3 Total emissions by time of the day

```{r, fig.height=5, fig.width=8}
emi_list$time_column <- emi_list$tp_model$timestamp  #### ja nao podia ser feito dentro da funcao?

emi_by_time <- emis_summary(emi_list = emi_list,
                                emi_var = "emi",
                                by = "time",
                                time_column = "time_column",
                                veh_var = "veh_type",
                                pol_var = "pollutant") 
head(emi_by_time) 

# plot
ggplot(data = emi_by_time) +
  geom_col(aes(x = factor(time_column), y = as.numeric(emi/1000), fill = pollutant),
           color=NA, show.legend = FALSE) +
  labs(y="Total emissions (Kg)", x="Hour of the day") +
  facet_wrap(~pollutant, scales = "free", nrow = 2) + 
  theme_minimal()

```


## 6.4 Spatial distribution of emissions

666666666 by road segment ??????????????


Finally, users can analyze how public transport emissions are spatially distributed. To do this, the `emis_grid()` function helps aggregate emission estimates over any custom spatial vector data (`sf POLYGON`). 

Let's create a regular hexagonal grid for this example.

```{r, message = FALSE}
# create spatial grid
mygrid <- sf::st_make_grid(
  x = sf::st_bbox(emi_list$tp_model$geometry)
  , cellsize = 0.25 / 200
  , crs= 4329
  , what = "polygons"
  , square = FALSE)

mygrid <- sf::st_sf(data.frame(id=1:length(mygrid), geom=mygrid))
plot(mygrid)
```

```{r, eval=TRUE, message=FALSE, fig.width=5, fig.height=5}
# we add info of emissions into gps file
emi_list$tp_model <- cbind(emi_list$tp_model, emi_list$emi) #### ja nao podia ser feito dentro da funcao?


mygrid_emi <- emis_grid(data = emi_list$tp_model,
                        emi = "CO2_Euro_III",    #### TOdo CO2 e Todo PM10 ?
                        grid = mygrid,
                        time_class = 'all periods')

ggplot() +
  geom_sf(data = mygrid_emi, aes(fill= as.numeric(CO2_Euro_III)/1000), color=NA) +
  scale_fill_continuous(type = "viridis")+
  labs(fill = "CO2 (Kg)")+
  theme_void()


```




## Credits

The `gtfs2gps` package is developed by a team at the Institute for Applied Economic Research (Ipea) with collaboration from the National Institute for Space Research (INPE), both from Brazil. If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/rafapereirabr/gtfs2emis).