---
title: 'Introduction to gtfs2emis'
author: "Joao Bazzo, Rafael H. M. Pereira, Pedro R. Andrade"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "`gtfs2emis` is an R package to estimate public transport emissions based on GTFS data. The package allows users to estimate the emission levels of several types of pollutants at high spatial and temporal resolutions simply using a GTFS feed and some information on fleet characteristics."
urlcolor: blue
vignette: |
  %\VignetteIndexEntry{Introduction to gtfs2emis} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

# Introduction

`gtfs2emis` is an R package to estimate hot exhaust emissions of public transport systems based on GTFS data. The package allows users to estimate the emission levels of several types of pollutants at high spatial and temporal resolutions simply using a GTFS feed and some information on fleet characteristics. To estimate the emission levels from a given public transport system, users need to:
1. Input a GTFS.zip file
2. Input data.table with a few characteristics of the public transport fleet (such as age and fuel)
3. Select an emission factor model from the models provided by the package, which currently include models from the US, Europe and Brazil.
4. Select which pollutants should be estimated from a list with of over 15 pollutants provided by the package.

This vignette introduces the main functions of the `gtfs2emis` package and shows a step-by-step reproducible example of how to use the package.


# Installation

One can install `gtfs2emis` from CRAN (recommended) or the development version from Github:

```{r, eval = FALSE, message = FALSE}
# From CRAN
install.packages("gtfs2emis")

# Dev. version with latest features
utils::remove.packages('gtfs2emis')
devtools::install_github("ipeaGIT/gtfs2emis")

```

# Package overview

Before we start, let's load a few packages we'll be using in this vignette:

```{r, message = FALSE}
library(gtfs2emis)
library(gtfstools)
library(gtfs2gps)
library(data.table)
library(magrittr)
library(ggplot2)
library(units)
library(sf)
```


The `gtfs2mis` has a few core functions:

* Estimate public transport emissions
  - `transport_model()` to convert GTFS data into GPS-like `data.table` with the space-time positions and speeds of public transport vehicles.
  - `emission_model()` to estimate hot-exhaust emissions based on (input from the transport model and data on fleet characteristics passed by the user.
  
* Post-process / analyze the data

- `emis_summary()` to aggregate emission estimates by time, vehicle type or road segment.
- `emis_grid()` to spatially aggregate emission estimates using any custom spatial grid or polygons.



# Demonstration on sample data

To demonstrate how the `gtfs2emis` package works, we will be using a small sample data for the city of Porto Alegre , Brazil. In this example, we'll be estimating CO2 emissions of bus services during business days.

Let's load the data.
<!-- The first step selects the GTFS and the desired route to model emissions. Here we use one example of `shape_id == "176-1"` in Porto Alegre (Brazil). Detailed functions of GTFS editing can be found in `gtfs2gps` and `gtfstools` packages. -->

```{r, message = FALSE}
gtfs <- gtfs2gps::read_gtfs(system.file("extdata/poa.zip", package = "gtfs2gps")) %>%
        # gtfstools::filter_by_weekday(weekday ='monday') 
        gtfs2gps::filter_week_days() 
        # %>% gtfstools::filter_by_shape_id(shape_id = '176-1') 

```


## 1. Transport model

The first step is to generate the transport model using the `transport_model()` function. This function converts GTFS data into a GPS-like data.table, fixing extreme or missing values of speed and transforming the output into a sf-linestring, which is the required input for emissions estimates. The user can input either a string with the file path where the a gtfs.zip file is stored, or an object of class `"gtfs" "list"`, generated with `gtfs2gps::read_gtfs()`.

### [666] Se a ideia do transport model for permitir fazer correcoes de velocidade, entao a gente precisa expor para os usuarios as mesmas funcoes da  função `gtfs2gps::adjust_speed()`.

```{r, message = FALSE}
tp_model <- transport_model(gtfs_data = gtfs, parallel = TRUE)

```

Here is how the output of the transport model looks like. In essence, it's a trajectory data.table with the space-time position and speed of trips segments for each single veihcle of the public transport system.


```{r, message = FALSE, fig.width=3, fig.height=3}
head(tp_model)

tp_model <- sf::st_as_sf(tp_model) #[666] Essa transfomacao pra sf nao devia ser feita dentro do transport_model() ?
plot(tp_model["speed"])
```


## 2. Fleet data

The next step is to prepare a `data.frame` with some characteristics of the public transport fleet. This can be either:
 - A detailed table that (1) brings info on the characteristics of each vehicle and, (2) tells the probability with which each vehicle type is allocated to each transport route.
 - A simple table with the overall composition of the fleet. In this case, the `gtfs2emis` will assume that fleet is homogeneously distributed across all routes.

We'll be working with a simple table on this example. For an example with detailed data, see the advanced use vignette of the package.

Now, the fleet data needs to be organized according to the data requirements of the emission factor model the user wants to consider. For instance, if a user wants to estimate emissions based on typical European fleet, the fleet data must include certain columns with the fleet characteristics that are used in the Emep-EEA methodology: vehicle type, Euro standard, technology and fuel. To check which columns and sets of vehicle characteristics are required by a given emission factor model, the user can read the documentation of the emission factor functions:

|Emission factor function| Country | Source | Type of buses                | Other required characteristics |
|------------------------|---------|--------|------------------------------|--------------------------------|
|`ef_cetesb_brazil()`    |Brazil | CETESB        | Micro, Standard, Articulated | Age, Fuel, EURO stage                    |
|`ef_emep_europe()`      |Europe | EMEP/EEA      | Micro, Standard, Articulated | Fuel, EURO stage, technology, load, slope|
|`ef_moves_usa()`        |US     | EMFAC         | Urban Buses                  | Age, Fuel                                |
|`ef_emfac_usa()`        |US     | MOVES U.S EPA | Urban Buses                  | Age, Fuel                                |
** [666] acho q o nome das funcoes ficariam melhor se seguissem o padrao `ef_pais_fonte`. Tipo `ef_brazil_cetesb()`, `ef_usa_moves()`, `ef_usa_emfac()`. Isso facilita pro usuario usar a funcao de autocomplete do rstudio quando estiver escrecendo o codigo**

Here are a few examples of data.frames with the fleet characteristics required by different emission factor models. Note that the table includes a column `fleet_composition`. This column tells what proportion of the fleet is represented by vehicles with each characteristic.

**Europe: EMEP/EEA**

```{r, message = FALSE}
fleet_data_ef_europe <- data.frame(  veh_type = c("Ubus Midi <=15 t","Ubus Std 15 - 18 t" ,"Ubus Artic >18 t")
                                   , euro = c("III","IV","V")
                                   , fuel = rep("D",3)
                                   , tech = c("-","SCR","SCR") #[666] o q significa o tracinho? nao tem info de tech daquele veiculo? nesse caso, o q o modelo faz?
                                   , fleet_composition = c(0.4,0.5,0.1)) #
fleet_data_ef_europe
```


**United States: EMFAC - California Air Resources Board**

```{r, message = FALSE}
fleet_data_ef_emfac <- data.frame(  veh_type = "BUS_URBAN_D"
                                  , model_year = 2011:2015
                                  , fuel = "D"
                                  , calendar_year = 2019
                                  , fleet_composition = rep(0.2,5))
fleet_data_ef_emfac
```

United States: Motor Vehicle Emission Simulator (MOVES) from Environmental Protection Agency:
```{r, message = FALSE}
fleet_data_ef_moves <- data.frame(  veh_type = "BUS_URBAN_D"
                                  , model_year = 2011:2015
                                  , fuel = "D"
                                  , calendar_year = 2016
                                  , fleet_composition = rep(0.2,5))
fleet_data_ef_moves
```

Brazil: Environmental Company of São Paulo (CETESB):
```{r, message = FALSE}
fleet_data_ef_cetesb <- data.frame( veh_type = c("BUS_MICRO_D", "BUS_URBAN_D", "BUS_ARTIC_D")
                                  , model_year = c(2010, 2012, 2018)
                                  , fuel = rep("D", 3)
                                  , fleet_composition = c(0.4, 0.4, 0.2))
fleet_data_ef_cetesb
```


## 3. Emission model

Finally, we use the `emission_model()` function to estimate hot exhaust emissions of our public transport system. Here, the user needs to pass the results from `transport_model()`, some fleet data as described above, and select which emission factor model and pollutants should be considered. The `emission_model()` generates a list with several files ?[666 files? ou vectors?] with emission estimates, storing information such as vehicle variables (`fuel`, `age`, `tech`, `euro`, `fleet_composition`), travel variables (`slope`, `load`, `gps`) or pollution (`EF`, `emi`). 

**[666] o modelos ainda nao calcula slopes, neh? sera q vale a pena manter isso no output? Pq se sim, entao tem q ter alguma explicacao na documentacao de qual valor esta sendo utilizado**

```{r, message = FALSE}
sf_emis <- emission_model( gps = tp_model
                          , ef_data_base = "emep_europe"
                          , fleet_data = fleet_data_ef_europe
                          , pollutant = c("CO","PM10","CO2","CH4","NOx")
                          )
class(sf_emis)
names(sf_emis)
```

## 4. Data post-processing

As the results are stored in a long list, gtfs2emis package has several function to post-process the results. `emis_summary` summarizes the emissions according to the variable of interest, such as vehicle type `veh_type`, `pollutant`, or `time`. In the last case, we need to add a `time_column` in the list, as this information is not returned by `emission_model()`.

```{r, message = FALSE}
sf_emis$time_column <- sf_emis$gps$timestamp

names(sf_emis)
my_emis_summary <- emis_summary(emi_list = sf_emis,
                    emi_var = "emi", 
                    by = "time", 
                    time_column = "time_column",
                    veh_var = "veh_type", 
                    pol_var = "pollutant") 
my_emis_summary
```
To visualize results by vehicle type `veh_type`:
```{r, message = FALSE}
my_emis_summary <- emis_summary(emi_list = sf_emis,
                                 emi_var = "emi", 
                                 by = "veh_type", 
                                 time_column = "time_column",
                                 veh_var = "veh_type", 
                                 pol_var = "pollutant") 
my_emis_summary
```
If available in `names(sf_emis)`, it is possible to add extra argument about fleet.
```{r, message = FALSE}
my_emis_summary <- emis_summary(emi_list = sf_emis,
                                emi_var = "emi", 
                                by = "veh_type", 
                                time_column = "time_column",
                                veh_var = c("veh_type", "euro", "fuel"),
                                pol_var = "pollutant") 
my_emis_summary
```
Checking the results by pollutant:
```{r, message = FALSE}
my_emis_summary <- emis_summary(emi_list = sf_emis,
                                 emi_var = "emi", 
                                 by = "pollutant", 
                                 time_column = "time_column",
                                 veh_var = "veh_type", 
                                 pol_var = "pollutant") 
my_emis_summary
```
If the user wants to have all results in a data.table format:
```{r, message = FALSE}
sf_emis$time_column <- sf_emis$gps$timestamp
my_emis_dt <- emi_to_dt(emi_list = sf_emis
                    ,emi_vars = "emi"
                    ,veh_vars = c("veh_type", "euro", "fuel")
                    ,pol_vars = "pollutant"
                    ,segment_vars = "time_column")
head(my_emis_dt, 5)
```

To visualize emissions spatially in a grid, the user nees to provide info about the desired grid. An operation of intersection can be done using `emis_grid()`. Here, we use `vein::make_grid()` to produce a basic grid.

```{r, message = FALSE}
grid_gps <- vein::make_grid(spobj = sf_emis$gps, width =  0.25 / 102.47) # 500 meters
grid_gps <- sf::st_as_sf(grid_gps)
```

```{r, message = FALSE, fig.width=4,fig.height=4}
names(sf_emis$emi)
# we add info of emissions into gps file
sf_emis$gps <- cbind(sf_emis$gps,sf_emis$emi)

pol_grid <- emis_grid(data = sf_emis$gps,
                     emi = "CO_Euro_IV",
                     grid = grid_gps,
                     time_class = 'all periods')
plot(pol_grid["CO_Euro_IV"])
```







 
## 4. Emissions estimates

As the emissions varies with `shape_id` and `fleet`, we compute emissions separately, in a loop.

```{r}
linestring_poa$shape_id %>% unique()
```

```{r, message = FALSE}
emi_data <- lapply(unique(gps_poa$shape_id),function(i){
  tmp_ef_euro <- ef_emep_europe(speed = linestring_poa[shape_id == i]$speed,
                                ,veh_type = fleet_data_ef_europe[shape_id == i]$veh_type
                                ,euro =  fleet_data_ef_europe[shape_id == i]$euro
                                ,pollutant = c("CO2","NOx")
                                ,fuel =  fleet_data_ef_europe[shape_id == i]$fuel
                                ,tech =  fleet_data_ef_europe[shape_id == i]$tech
                                ,as_list = TRUE)
  tmp_emis <- emis(fleet_composition = fleet_data_ef_europe[shape_id == i]$fleet_composition
                   ,dist = units::set_units(linestring_poa[shape_id == i]$dist,"km")
                   ,ef = tmp_ef_euro
                   ,aggregate = FALSE
                   ,as_list = TRUE)
  return(tmp_emis)
})
```

## 5. Emission into grid

We now show the example for `CO2` emissions.

```{r}
# Create Grid
grid_gps <- vein::make_grid(spobj = sf::st_as_sf(linestring_poa$geometry)
                            , width =  0.25 / 102.47) # 500 meters
grid_gps <- sf::st_as_sf(grid_gps)

tmp_emis_dt <- lapply(1:length(emi_data),function(i){ # i = 3
  tmp_emi_dt <- emi_to_dt(emi_list = emi_data[[i]]
                          ,emi_vars = "emi"
                          ,veh_vars = "veh_type"
                          ,pol_vars = "pollutant")
  tmp_gps <- cbind(linestring_poa[shape_id == unique(gps_poa$shape_id)[i]]
                   ,tmp_emi_dt[pollutant == "CO2"])
  tmp_gps <- sf::st_as_sf(tmp_gps)
  tmp_emi_grid <- emis_grid(data = tmp_gps
                            ,emi = "emi"
                            ,grid = grid_gps)
  return(tmp_emi_grid)
}) %>% data.table::rbindlist()
```

Summing all emissions and plotting:

```{r, fig.width=4,fig.height=4}
tmp_emis_dt <- tmp_emis_dt[,total_emi := sum(emi), by = id][, .SD[1], by = id]
tmp_emis_dt2 <- sf::st_as_sf(tmp_emis_dt)

plot(tmp_emis_dt2["emi"])
```

