---
title: "Exploring Non-Exhaust Emission Factors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Exploring Non-Exhaust Emission Factors} 
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown} 
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

When assessing vehicle emissions inventories for particles, one relevant step is taking into account the non-exhaust processes, such as tire, brake and road wear. The `gtfs2emis` incorporates the non-exhaust emissions methods from [EMEP-EEA](https://www.eea.europa.eu//publications/emep-eea-guidebook-2019).
 
The following equation is employed to evaluate emissions originating
from tire and brake wear

$$
 TE_i = dist \times EF_{tsp}(j) \times mf_s(i) \times SC(speed)
$$ where:

-   $TE(i)$ = total emissions of pollutant i (g),
-   $dist$ = distance driven by each vehicle (km),
-   $EF_{tsp}(j)$ = TSP mass emission factor for vehicles of category j
    (g/km),
-   $mf_s(i)$ = mass fraction of TSP that can be attributed to particle
    size class i,
-   $SC(speed)$ = correction factor for a mean vehicle travelling at a
    given speed (-).

## Tire

In the case of heavy-duty vehicles, the emission factor needs the
incorporation of vehicle size, as determined by the number of axles, and
load. These parameters are introduced into the equation as follows:

$$EFTSP^{hdv}_{tire} = 0.5 \times N_{axle} \times LCF_{tire} \times EFTSP^{pc}_{tire}$$

where:

-   $EFTSP^{hdv}_{tire}$ = TSP emission factor for tire wear from
heavy-duty vehicles (g/km),
-   $N_{axle}$ = number of vehicle axles (-),
-   $LCF_{tire}$ = a load correction factor for tire wear (-),
-   $EFTSP^{pc}_{tire}$ = TSP emission factor for tire wear from passenger
car vehicles (g/km).

and $$LCF_{tire} = 1.41 + (1.38 \times LF)$$

where:

-   $LF$ = load factor (-), ranging from 0 for an empty bus to 1 for a fully laden one.

The function considers the following look-up table for number of vehicle
axes:

| vehicle class (j)    | number of axes |
|----------------------|----------------|
| Ubus Midi \<=15 t    | 2              |
| Ubus Std 15 - 18 t   | 2              |
| Ubus Artic \>18 t    | 3              |
| Coaches Std \<=18 t  | 2              |
| Coaches Artic \>18 t | 3              |

The size distribution of tire wear particles are given by:

| particle size class (i) | mass fraction of TSP |
|-------------------------|----------------------|
| TSP                     | 1.000                |
| PM10                    | 0.600                |
| PM2.5                   | 0.420                |
| PM1.0                   | 0.060                |
| PM0.1                   | 0.048                |

Finally, the speed correction is:

-   $SC_{tire}(speed) = 1.39$, when $speed < 40 km/h$;
-   $SC_{tire}(speed) = -0.00974 \times speed + 1.78$, when $40 <= speed <= 90 km/h$;
-   $SC_{tire}(speed) = 0.902$, when $speed > 90 km/h$.

```{r}
library(gtfs2emis)
emi_europe_emep_wear(dist = units::set_units(1,"km"),
                              speed =  units::set_units(30,"km/h"),
                              pollutant = c("PM10","TSP","PM2.5"),
                              veh_type = "Ubus Std 15 - 18 t",
                              fleet_composition = 1,
                              load = 0.5,
                              process = c("tyre"),
                              as_list = TRUE)
```

## Brake

The heavy-duty vehicle emission factor is derived by modifying the
passenger car emission factor to conform to experimental data obtained
from heavy-duty vehicles.

$$EFTSP^{hdv}_{brake} = 1.956 \times LCF_{brake} \times EFTSP^{pc}_{brake}$$

where:

-   $EFTSP^{hdv}_{brake}$ = heavy-duty vehicle emission factor for TSP,
-   $LCF_{brake}$ = load correction factor for brake wear,
-   $EFTSP^{pc}_{brake}$ = passenger car emission factor for TSP.


$$LCF_{brake} = 1 + (0.79 \times LF),$$

where:

-  $LF$ = load factor (-), ranging from 0 for an empty bus to 1 for a fully laden one.

The size distribution of brake wear particles are given by:

| particle size class (i) | mass fraction of TSP |
|-------------------------|----------------------|
| TSP                     | 1.000                |
| PM10                    | 0.980                |
| PM2.5                   | 0.390                |
| PM1.0                   | 0.100                |
| PM0.1                   | 0.080                |

Finally, the speed correction is:

-   $SC_{brake}(speed) = 1.67$, when $speed < 40 km/h$;
-   $SC_{brake}(speed) = -0.0270 \times speed + 2.75$, when $40 <= speed <= 95 km/h$;
-   $SC_{brake}(speed) = 0.185$, when $speed > 95 km/h$.

```{r, message = FALSE}
emi_europe_emep_wear(dist = units::set_units(1,"km"),
                              speed =  units::set_units(30,"km/h"),
                              pollutant = c("PM10","TSP","PM2.5"),
                              veh_type = "Ubus Std 15 - 18 t",
                              fleet_composition = 1,
                              load = 0.5,
                              process = c("brake"),
                              as_list = TRUE)
```

## Road Wear

Emissions are calculated according to the equation:

$$TE(i) = dist \times EF^{road}_{tsp}(j) \times mf_{road}$$

where:

-   $TE(i)$ = total emissions of pollutant i (g),
-   $dist$ = total distance driven by vehicles in category j (km),
-   $EF^{road}_{tsp}$ = TSP mass emission factor from road wear for vehicles j (0.0760 g/km),
-   $mf_{road}$ = mass fraction of TSP that can be attributed to particle size class i (-).

The following table shows the size distribution of road surface wear
particles

| particle size class (i) | mass fraction of TSP |
|-------------------------|----------------------|
| TSP                     | 1.00                 |
| PM10                    | 0.50                 |
| PM2.5                   | 0.27                 |

```{r, message = FALSE}
emi_europe_emep_wear(dist = units::set_units(1,"km"),
                              speed =  units::set_units(30,"km/h"),
                              pollutant = c("PM10","TSP","PM2.5"),
                              veh_type = "Ubus Std 15 - 18 t",
                              fleet_composition = 1,
                              load = 0.5,
                              process = c("road"),
                              as_list = TRUE)
```

## Viewing Emissions

Users can also use one single function to apply for more than one process (e.g. tire, brake and road), as shown below.

```{r, message = FALSE,fig.height=3, fig.width=8}
library(units)
library(ggplot2)

emis_list <- emi_europe_emep_wear(dist = units::set_units(rep(1,100),"km"),
                     speed =  units::set_units(1:100,"km/h"),
                     pollutant = c("PM10","TSP","PM2.5"),
                     veh_type = c("Ubus Std 15 - 18 t"),
                     fleet_composition = c(1),
                     load = 0.5,
                     process = c("brake","tyre","road"),
                     as_list = TRUE)
ef_dt <- gtfs2emis::emis_to_dt(emis_list,emi_vars = "emi"
                               ,segment_vars = "speed")
ggplot(ef_dt)+
  geom_line(aes(x = as.numeric(speed),y = as.numeric(emi),color = pollutant))+
  facet_wrap(facets = vars(process))+
  labs(x = "Speed (km/h)",y = "Emissions (g)")+
  theme_minimal()
```
### References

EMEP/EEA data and reports can be accessed in the following links:
- 2019 edition \url{https://www.eea.europa.eu/publications/emep-eea-guidebook-2019/part-b-sectoral-guidance-chapters/1-energy/1-a-combustion/1-a-3-b-vi/view}.


## Report a bug

If you have any suggestions or want to report an error, please visit
[the package GitHub page](https://github.com/ipeaGIT/gtfs2emis/issues).
